3.5: Causal Inference and DAGs

### 2019-11-07 content/r/17-r-practice.Rmd
Final fixes to practice 17 content, slides 17

Note what we are hypothesizing:
  - `bwght` is caused by `cigs` and `inc`
  - `cigs` are caused by `price`, `educ`, and `inc`
  - `inc` is caused by `educ`
See also how this is written into the notation in R to make the DAG.

>>>>

Note what we are hypothesizing:

- `bwght` is caused by `cigs` and `inc`
- `cigs` are caused by `price`, `educ`, and `inc`
- `inc` is caused by `educ`

See also how this is written into the notation in R to make the DAG.

### 2019-11-07 static/slides/17-slides.Rmd
Final fixes to practice 17 content, slides 17

1. $discr \rightarrow wage$ (Causal path)
2. $discr \rightarrow occup \rightarrow wage$ (Causal path)
3. $discr \leftarrow gender \rightarrow occup \rightarrow wage$ (Open back door)
4. $discr \leftarrow gender \rightarrow occup \leftarrow abil \rightarrow wage$ (BLOCKED back door!)
5. $discr \rightarrow occup \leftarrow abil \rightarrow wage$ (BLOCKED back door!)

@@

- If we *control* for occupation:
  - opens backdoor paths 4 & 5!
  - creates spurious `cor(abil,discr)` where there wasn't one!
  - also closes causal path 2! (discriminatory reasons genders choose different occupations)

- **Can't identify causal effect** controlling for `occup`!
  - *Could* if you could control for `occup` *and* `abil` (but perhaps `abil` is unobserved)

- Research on this needs to get clever!

@@

# The Front Door Criterion: Mediators

@@

@@

# The Front Door Criterion: Mediators

@@

Thus, to achieve identification, control for the minimal amount of variables such that:

@@

# What are DAGs Good For?

- Omitted variable bias

- Help visualize how later methods (difference-in-difference, fixed effects, regression discontinuity, instrumental variables) help identify causal effectsp

>>>>

1. $discr \rightarrow wage$ (Causal path)
2. $discr \rightarrow occup \rightarrow wage$ (Causal path)
3. $discr \leftarrow gender \rightarrow occup \rightarrow wage$ (BLOCKED back door!)
4. $discr \leftarrow gender \rightarrow occup \leftarrow abil \rightarrow wage$ (BLOCKED back door!)
5. $discr \rightarrow occup \leftarrow abil \rightarrow wage$ (BLOCKED back door!)

@@

- If we *control* for occupation:
  - opens backdoor paths 3, 4, 5!
  - creates spurious `cor(abil,discr)` where there wasn't one!
  - also closes causal path 2! (discriminatory reasons genders choose different occupations)

- **Can't identify causal effect** controlling for `occup` alone!
  - *Could* if you could control for `occup` *and* `abil` (but perhaps `abil` is unobserved)

@@

# The Front Door Criterion: Mediators I

@@

# The Front Door Criterion: Mediators II

- If we control for $M$, this would block the front-door!

@@

# The Front Door Criterion: Mediators III

@@

Thus, to achieve .onfire[causal identification], control for the minimal amount of variables such that:

@@
